---
title: "Exploratory Data Analysis of White wine quality"
Date: September 25, 2016
output: pdf_document
Author: Naga Ramesh Kamisetti
---
This is my fourth project in the nano degree program of Data Analysis by 
Udacity. In this project I leveraged the R libraries to explore white wine 
quality dataset available [here](https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/)
and the description of the variables is provided [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt)

This dataset consists of physicochemical properties of winesamples collected to 
understand how these properties affect quality of the wine. Quality, in this 
dataset is a sensory preferance given by expert wine tasters on a sclae of 0-10 
where 0 indicates "bad" and 10 indicates "good".

This project is organized to include my thought process as I analyze different 
physicochemical properties, their relationships and their affect on sensory 
based quality. I included univariate, bivariate, multivariate analysis along 
with different visualizations. 

```{r include=FALSE, message=FALSE, warning=FALSE, echo=FALSE, install_Packages}
# Installing all the packages required to run the code in this document.
install.packages("ggplot2", repos="http://cran.rstudio.com/")
install.packages("gridExtra", repos="http://cran.rstudio.com/")
install.packages("GGally", repos="http://cran.rstudio.com/")
install.packages("corrplot", repos="http://cran.rstudio.com/")
install.packages("e1071", repos="http://cran.rstudio.com/" )

```

```{r message=FALSE, warning=FALSE,echo=FALSE, loading_packages}
# loading all required libraries in to the workspace. 
library(ggplot2)
library(gridExtra)
library(GGally)
library(corrplot)
library(e1071)
```

Loading the data set in to workspace. 
```{r message = FALSE, warning=FALSE, loading_dataset}
# Load the Data
winedata=read.csv("wineQualityWhites.csv")

```

After loading the data set, I want to see the outline of the data set. For this 
I will use summary and dim functions.

```{r message=FALSE, warning= FALSE,summary }
# to get the dimensions of the data set
dim(winedata)

#to get the names of the variables
names(winedata)

#structure of the data set
str(winedata)

# to get the summary of all variables
summary(winedata)
```

As seen from the above result, this data set has 4898 rows and 13 variables. 
From the summary it is observed that the first variable "x" is used as an index 
and it is not required for the analysis. So, I will delete this variable before 
proceeding to further analysis. 

The variable names are very descriptive and explains what physicochemical 
property they represent. The maximum value for almost all variables is very high
compared to the median value indicating the presence of outliers. I want to plot
histograms to see if my  intuition is correct. Plotting histograms also explains
the distribution of the variable.

Quality is taken as a quantittative variable as indicated by mean instead of 
qulitative value. I will use it to craete a new quality_factor variable as a
ordered factor. Residual sugar, and the two types of sulfur dioxide have high 
variablity as can be seen from their minimum and maximum values.Citric acid has 
a minimum value of 0. The pH values are all lesthan 4 indicating wine is acidic.

```{r message=FALSE, warning= FALSE, clean_dataset}
# deleting "x" variable
winedata<-winedata[,-1]

# creating a ordered factor variable representing quality
winedata$quality_factor <- factor(winedata$quality, ordered = TRUE)

# looking at the summary afteer the intial modifications
summary(winedata)
```

As seen from the summary of quality_factor variable there are very less samples 
representing quality levels of 3 and 9. Also, quality level of 6 is most 
represented in this sample dataset. There are no samples representing quality 
levels of 0,1,2 and best quality 10.I believe that the analysis of this data can
be improved by grouping quality levels as "bad", "average", and "good". This can
be done becuase there is no specific cut for each quality level. They all are 
sensory perception results by wine tasters. 

```{r message=FALSE, warning= FALSE, rating_variable}
# attaching the dataset here so that code can be readable
attach(winedata)
# creating a quality rating variable
winedata$rating <- factor(
                   ifelse((quality == '3'| quality == '4'),"bad",
                    ifelse((quality == '5'| quality== '6'),"average","good")),
                   levels=c('bad','average','good')
                   )               

```

For every sample having a quality value of 3 or 4 I grouped them as having a 
rating of "bad" and for samples of quality 5 or 6, I grouped them as having a 
rating of "average", and for all other samples having quality 7,8,9 as "good".

```{r message=FALSE, warning= FALSE, summary_after_datacleaning}
summary(winedata)
```

There are 1060 observations having a rating of "good" and 183 having rating 
"bad" and the rest having a rating "average". 

# Univariate Analysis

I want to see how the ariables are distributed. Whenever there is an indication
of outliers from the summary, I used limits to exclude some of the outliers. 

```{r message=FALSE, warning= FALSE,quality_histogram}
p1 = ggplot(winedata, aes(x=quality))+
     geom_bar()
p2 = ggplot(winedata, aes(x=rating,fill = rating))+
     geom_bar()
grid.arrange(p1,p2,ncol=2)
```

As seen from the plot on the left, the number of observations falling under 
higher quality is less. Hence my new rating variable is created. This new
rating variable distribution is shown on the right. Most of the samples fall 
under average quality.

```{r message=FALSE, warning= FALSE,FA_histogram}
ggplot(winedata, aes(x=fixed.acidity))+
  xlim(quantile(winedata$fixed.acidity,0.01),
       quantile(winedata$fixed.acidity,0.99))+
  geom_histogram(binwidth = 0.1)
```

From the above graph it can be clearly identified that the fixed acidity 
vriable is approximately normally distributed. This fact can be used when 
modelling for quality using this variable. 

```{r message=FALSE, warning= FALSE,VA_histogram}
ggplot(winedata,aes(x=volatile.acidity))+
  xlim(quantile(winedata$volatile.acidity,0.01),
       quantile(winedata$volatile.acidity,0.99))+
  geom_histogram(binwidth = 0.01)
```

It can be seen that volatile acidity distribution is right skewed. I want to see
if a transformation of this variable makes it any near to normal distribution.

```{r message=FALSE, warning= FALSE,VA_log_histogram}
ggplot(winedata,aes(x=volatile.acidity))+
  geom_histogram()+
  scale_x_log10()
```

As can be seen, this log transform of volatile acidity is approximately normal.
I will use this transformation whenever I need to use volatile acidity.

```{r message=FALSE, warning= FALSE,ca_histogram}
ggplot(winedata,aes(x=citric.acid))+
  xlim(quantile(winedata$citric.acid,0.01),
       quantile(winedata$citric.acid,0.99))+
  geom_histogram(binwidth = 0.01)
```

Citric acid is approximately normal after excluding outliers.

```{r message=FALSE, warning= FALSE,RS_histogram}
ggplot(winedata,aes(x=residual.sugar))+
  xlim(quantile(winedata$residual.sugar,0.01),
       quantile(winedata$residual.sugar,0.99))+
  geom_histogram(binwidth = 0.1)
```

Residual sugar distribution is very different from normal distribution. 
I would like to use a log transform and see if the distribution changes.

```{r message=FALSE, warning= FALSE,RS_log_histogram}
ggplot(winedata,aes(x=residual.sugar))+
  scale_x_log10()+
  geom_histogram()
```

Even this log transform is not near to normal, but it looks like this log 
transform has bi modal distribution. This distribution seems better compared 
to the untransformed variable. I will use this in my further analysis.

```{r message=FALSE, warning= FALSE,cl_histogram}
ggplot(winedata,aes(x=chlorides))+
  xlim(quantile(winedata$chlorides,0.01),
       quantile(winedata$chlorides,0.96))+
  geom_histogram(binwidth = 0.005)
```

Most of the chloride content range from 0.009 to 0.05 g/dm^3. There are some 
outliers which I excluded in this graph. If I consider this variable for further
analysis then I will include all values of chlorides to see its affect on 
wine quality. 

```{r message=FALSE, warning= FALSE,sulfur_histogram}
a1 = ggplot(winedata,aes(x=free.sulfur.dioxide))+
     xlim(quantile(winedata$free.sulfur.dioxide,0.01),
       quantile(winedata$free.sulfur.dioxide,0.99))+
     geom_histogram(binwidth = 1)
a2 = ggplot(winedata,aes(x=total.sulfur.dioxide))+
     xlim(quantile(winedata$total.sulfur.dioxide,0.01),
       quantile(winedata$total.sulfur.dioxide,0.99))+
     geom_histogram(binwidth = 1)
grid.arrange(a1,a2, ncol=2)
```

The two forms sulfur dioxide have normal distribution.

```{r message=FALSE, warning= FALSE,density_histogram}
ggplot(winedata,aes(x=density))+
       xlim(quantile(winedata$density,0.001),
         quantile(winedata$density,0.999))+
       geom_histogram(binwidth = 0.00065)
```

The density values are appoximately closer to 1, density of water. 

```{r message=FALSE, warning= FALSE,densityr_histogram }
ggplot(winedata,aes(x=sulphates))+
       xlim(quantile(winedata$sulphates,0.01),
         quantile(winedata$sulphates,0.99))+
       geom_histogram(binwidth = 0.01)
```

Sulphates have approximate normal distribution, but there is a small skewness 
towards right. Most of the values are less than 0.6

```{r message=FALSE, warning= FALSE,alcohol_histogram}
ggplot(winedata,aes(x=alcohol))+
       geom_histogram(binwidth = 0.1)
```

Alcohol content varies from 8 to 14 but most of the values range from 9-13.

```{r message=FALSE, warning= FALSE,pH_histogram}
ggplot(winedata,aes(x=pH))+
       xlim(quantile(winedata$pH,0.01),
         quantile(winedata$pH,0.99))+
       geom_histogram(binwidth = 0.05)
```

As already noted from the summary, all these pH values indicate that wine is 
acidic. most of the values range from 3.0 - 3.4 Now, I got an idea of how each 
individual variable is distributed.

Before proceeding further, because the number of high quality samples are very 
less, I would like to see all the variables corresponding to these high quality 
wine samples.

```{r message=FALSE, warning= FALSE, good_quality_wine}
winedata[(winedata$quality=='9'),]
```

For these high quality sample wines, most of the variables have values in the 
average range. But one strange thing identified from this analysis is that, the 
alcohol value for all these samples is more than 12 except for the row 775 
sample. But, the residual sugar value for this sample is very high, 10.6, 
ompared to the remaining four samples. This created a doubt for the existence
of a relationship between quality, alcohol content and residual sugar. In my 
further analysis, I will observe this association more closely to find out 
if this relationship does actually exist or it is just a outlier.

Before I attempt to create any model to predict wine quality, I want to see if 
any collinearity exists between these predictor variables. I strongly believe 
that some of the variables are correlated. Also, I want to findout the 
correlation between quality and these predictor variables. 

Multicollinearity is very problematic and its identification before modelling 
can help us in using better predictors in the model. 

# Bivariate and Multivariate Analysis

```{r message=FALSE, warning= FALSE, correlation}
corrplot.mixed(cor(winedata[,1:12]),tl.pos = "lt")
```

Correlation between quality and othe rvariables:

* Alcohol and quality have a correlation of 0.44
* Density and quality have a correlation of -0.31
* chlorides and quality have a correlation of -0.21

All other variables have either very less or no correlation with quality.

Correlation among other variables:

Residual Sugar and density have a correlation of 0.84. This is a very strong 
relationship. Alcohol and density have a strong negative correlation, -0.78. 
The two forms of sulfur dioxide have a correlation of 0.62, strong correlation 
as predicted. Density has a correlation of 0.53 with total sulfur dioxide. 
Alcohol has a negative correlation of -0.45 with both residual sugar and total 
sulfur dioxide. Alcohol and chlorides have a -0.36 correlation.

The most important observation is the negative correlation between residual 
sugar and alcohol, as I predicted for the existence of a relationship between 
these two variables above. This proved that the sample 775 observed above is 
not just a outlier. Mostly, alcohol and density are correlated with other 
variables  and also with quality indicating that these can be good predictors 
in a model for predicting quality.

As I found that correlation exists between some of the variables, I want to plot
them to better understand their relationship. From the above plot I understood 
that alcohol content, residual sugar, density, chlorides, sulfur dioxide affect 
quality either directly or becuase of their relationship with other variables. 
Now I will plot some of these relationships.

```{r message=FALSE, warning= FALSE, quality_al}
ggplot(winedata,aes(y=alcohol,x=quality_factor, fill = rating))+
  geom_boxplot()
```

From this plot it is very clear that alcohol is good predictor and quality 
improves with high alcohol content. But I want to test this intuition when other correlating avriables are present. Now I'll consider affect of residual sugar 
in the rpesence of alcohol content.

```{r message=FALSE, warning= FALSE, quality_rs_al}
ggplot(winedata,aes(x=alcohol, y=residual.sugar, color = rating))+
   ylim(quantile(winedata$residual.sugar,0.0),
       quantile(winedata$residual.sugar,0.99))+
  geom_point()+
  geom_vline(xintercept = 11.5, color = "red")
```

Most of the high quality wines have a alcohol content of above 11.5 as indicated
by the red vertical line. This plot is cumbersome with all quality of wines. So 
I decided to plot only good quality wines.

```{r message=FALSE, warning= FALSE, quality_rs_al2}
wine_good = winedata[(winedata$rating=="good"),]
ggplot(wine_good,aes(x=alcohol, y=residual.sugar))+
   ylim(quantile(wine_good$residual.sugar,0.0),
       quantile(wine_good$residual.sugar,0.99))+
  geom_point()+
  geom_vline(xintercept = 9, color = "red")+
  geom_hline(yintercept = 8, color = "blue")
```

When the alcohol content is further less, then samples having high residual 
sugar values, above blue line,  have good quality rating. From the above graph, 
I conclude that alcohol is very stron g predictor and residual sugar is 
important only when the alcohol content is less.

Now, I want to see how density affects the relationship of alcohol and quality 
given that alcohol and density have a negative correlation. 

```{r message=FALSE, warning= FALSE, quality_al_den}
ggplot(winedata,aes(x = alcohol, y = density, color = rating))+
  ylim(quantile(winedata$density,0.0),
       quantile(winedata$density,0.99))+
  geom_point()+
  geom_vline(xintercept = 11, color = "black")
```

Again, alcohol is dominating even in the presence of density. samples having 
alcohol content above 11 (black vertical line)  tends to have higher quality
value irrespective of the density value. Hence, as suggested by the correlatin 
plot, the affect of density is taken care by the alcohol content as a predictor 
for predicting quality.

Now, I want to see how total sulfur dioxide affects quality in the presence of 
alcohol content as it has a negative correlation with alcohol.

```{r message=FALSE, warning= FALSE, quality_al_tso2}
ggplot(winedata,aes(x = alcohol, y = total.sulfur.dioxide, color = rating))+
  geom_point()
```

Again, it is observed that alcohol dominates. From all these graphs I conclude 
that alcohol is very strong predictor. 

Now I want to see how chlorides affects wine quality.

```{r message=FALSE, warning= FALSE, quality_cl}
ggplot(winedata,aes(x = quality_factor, y = chlorides, fill = rating))+
geom_boxplot()
```

As I previously noted there are outliers in chloride distribution that can be 
seen in the bove diagram for quality levels of 5 and 6. 

```{r message=FALSE, warning= FALSE, quality_cl2}
ggplot(winedata,aes(x = quality_factor, y = chlorides, fill = rating))+
  ylim(quantile(winedata$chlorides,0.),
       quantile(winedata$chlorides,0.9))+
geom_boxplot()
```

Even though this plot suggests that quaity is more for lesser chloride content 
I suspect that this relationship must be further analyzed as the number of 
samples representing the quality level of 9 are just five. So I will plot a 
boxplot using rating variable. 

```{r message=FALSE, warning= FALSE, quality_cl3}
ggplot(winedata,aes(x = rating, y = chlorides, fill = rating))+
  ylim(quantile(winedata$chlorides,0.),
       quantile(winedata$chlorides,0.9))+
geom_boxplot()
```

After plotting with rating variable, it still shows that lesser chloride content 
results in sample of good quality. So, I conclude chloride content is also a 
good predictor for quality. But, I want to further test its affect in the 
presence of other variables. 

```{r message=FALSE, warning= FALSE, quality_cl_al}
ggplot(winedata,aes(x=alcohol,y=chlorides,color = rating))+
  geom_point()
```

In presence of alcohol, chlorides does not have any affect on quality as a 
predictor. This further strengthed my previous observation that alcohol is a 
strong predictor for quality.

# Model Building

I want to build a model to predict quality based on the observations from the 
above plots. First I want to perform a svm model using alcohol and density as 
predictors. 
 
```{r message=FALSE, warning= FALSE, model1 }
model1<-svm(rating ~ alcohol+density, winedata)
preds1<-predict(model1, winedata)

# finding accuracy
table(winedata$rating,preds1)
```

accuracy  = (3508+235)/(4898) = 0.7641

This is a good accuracy but, I used the same training data to find out accuracy.
So, it is not reliable but still is a good estimate. Now I want to check for the
affect of chloride as  predictor besides alcohol.

```{r message=FALSE, warning= FALSE, model2 }
model2<-svm(rating ~ alcohol+chlorides, winedata)
preds2<-predict(model2, winedata)

# finding accuracy
table(winedata$rating,preds2)
```

accuracy = (3496+245)/4898 = 0.7637

This is almost same as the previous one. This further strengthend my intuition 
that even though in th evisualization, chloride afected quality, over all it is 
not that effective.

```{r message=FALSE, warning= FALSE, model3 }
model3<-svm(rating ~ alcohol+total.sulfur.dioxide, winedata)
preds3<-predict(model3, winedata)

# finding accuracy
table(winedata$rating,preds3)
```

accuracy = (2+3515+216)/4898 = 0.76212 

Again this model did not improve my accuracy. One point to remember here is that
I am testing my models on the same data I used to train. This is because I don't
have enough data of quality values 9 so as to build a model with training data 
and test it with testing data.

Now, I want to test the model with only Alcohol, as all my previous analysis 
proved that quality is modtly affected by alcohol and in its presence as a 
predictor the other variables are not that effective. 

```{r message=FALSE, warning= FALSE, model4 }
model4<-svm(rating ~ alcohol, winedata)
preds4<-predict(model4, winedata)

# finding accuracy
table(winedata$rating,preds4)
```

accuracy = (3478+255) = 0.7621

This is interesting!! Just one predictor , alcohol content, produces an accuracy 
of almost same as previous models. This strengthend my orevious statements that
alchols supreeses the affect of other variables as predictors in its presence.

The best model so far is first model where I used alcohol along with density. 

# Final Plots

**Histogram of Wine Quality**

First I'll plot how quality is distributed in this dataset.

```{r message=FALSE, warning= FALSE, final_plot1}
a1 = ggplot(winedata,aes(x = as.factor(quality)))+
 geom_bar(color = I('black'),fill = I('#981133')) + 
  xlab("Wine Quality")+
  ylab("Number of Samples")+
  ggtitle("Histogram of Wine Quality")

a2 = ggplot(winedata, aes(x= rating,fill = rating))+
  geom_bar()+
  ggtitle(" Histogram of Wine Rating")+
  xlab("Wine Rating")+
  ylab("Number of Samples")
grid.arrange(a1,a2,ncol = 2)
```

This is the first and most important plot of this entire analysis. This plot 
shows that most of the samples have a quality level of 6. Also, very few samples represented quality level of 3 ,4, 8 and 9. The plot to the left helped me in 
creating the plot to the right where I created a rating variables taking 7,8, 
and 9 quality level as good, 3,4 quality level as bad and the rest as average. 
This new variable helped me when I performed multivariate analysis. 

**Quality VS Alcohol Content **

Now, I want to plot the relationship between quality and alcohol content. 

```{r message=FALSE, warning= FALSE, quality_al_finalplot2}
ggplot(winedata,aes(y=alcohol,x=quality_factor, fill = rating))+
  geom_boxplot()+
  ggtitle("Variation of Quality with Alcohol Content")+
  ylab(" Alcohol content (% by Vol)")+
  xlab("Quality level")
```

This plot is the heart of this analysis. This clearly shows the affect of 
alcohol on the quality level, which proved to be the most effective predictor. 
This plot shows that as the alcohol content is increasing, the quality level 
increased. This can be taken in an other way as all the wine tasters who 
performed this quality check liked alcohol. THis plot is the starting point for
my multivariate analysis as many other variables are correlated with alcohol. 

** Quality Level with Alcohol and Density**

Now I want to plot the affect of alcohol content and densoty together with 
quality as this combination proved to be the best model using SVM in my 
predictive model.

```{r message=FALSE, warning= FALSE, quality_al_finalplot3}
ggplot(winedata,aes(y=density,x=alcohol, color = rating))+
  geom_point()+
  ggtitle("Variation of Quality with Alcohol and Density")+
  ylim(quantile(winedata$densityr,0.01),
       quantile(winedata$density,0.95))+
  xlab(" Alcohol content (% by Vol)")+
  ylab("Density (g / cm^3)")
```

This plot showed that eventhough density is a good predictor, in presence of 
alcohol its affect is very less as samples having higher alcohol content tends 
to have high rating. But for samples having low alcohol content this might be a
good predictor, as my best model was obtained combining alcohol and density. 
Having density improved my accuracy slightly more cmpared to having only alchol
as predictor. 

# Summary and reflection

In this project I learne to analyze a large data set. Eventhough I didn't know 
anything about wine, I now understood that the quality of wine depends largely 
on alcohol content and density. I learned to use ggplot2 library and also 
reading documentation whenever required to plot good visualizations. 

I tarted with the summary of the data set and slowly built my analysis atarting 
with univariate analysis. This provided me with what to consider later in my 
analysis. I googled to find out abour corrplot library to draw the visual I 
used, as the base graph plot was very slow. I'm now confident to analyze further
more large data sets. 


# Further Improvements

In this analysis, I used same data for both training and testing my models. 
Someone can use better cross validation methods to find the true accuracy of the
models I built. This needs more attention because there are very less samples 
representing the quality levels of 3,4,8 and 9. 

In my models, the predicted classes never predicted quality level "bad" in most 
of the models. It would be interesting to know why this happend as my training 
data has a lot of samples representing "bad" quality rating. 

WHen I plotted alcohol affect vs quality I observed that quality level 0f 5 has 
an average alcohol content less than that of both quality levels 4 and 6. It 
would be interesting to know how thi happend given the fact that the general 
trend was to have high quality with increasing alcohol content. 

